Model Finding for (Co)recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-05 tex/main.tex
used 'admissible' terminology

%In this paper, we associate quantified formulas with uninterpreted functions (informally, those whose semantics they define).
In this paper, we consider \emph{annotated quantifed formulas} of the form
$\forallf{a} \vec x.\; \varphi$, where the annotation~$a$ is either
an uninterpreted function $\con{f} \in \sfuns{\Sigma}$ or the special symbol
$\fnull$.
Given $\con{f} : \tau_1 \times \ldots \times \tau_n \rightarrow \tau$,
an annotated
quantified formula $\forallf{\con{f}} \vec x.\; \varphi$ is a \emph{function definition}
(\,\emph{for $\con{f}$}\vthinspace) if $\vec x$ is a tuple of variables of type
$\tau_1, \ldots, \tau_n$ and $\forallf{\con{f}} \vec x.\; \varphi$ is equivalent to
$\forall \vec x.\;\, \con{f}( \vec x ) \teq t$ for some $t$. We will write $\forall \vec
x. \varphi$ as shorthand for $\forallf{\fnull} \vec x.\; \varphi$, and $\exists
\vec x. \varphi$ as shorthand for $\neg \forall \vec x.\; \neg \varphi$.

\begin{definition}
A formula $\varphi$ is \emph{well-defined with respect to functions} $\{ f_1, \ldots, f_n \}$ if it is of the form:
\begin{eqnarray} \label{eq:wdf}
(\forallf{f_1} \vec x_1.\; \varphi_1) \wedge \ldots \wedge (\forallf{f_n} \vec x_n.\; \varphi_n) \wedge \psi
\end{eqnarray}
where $f_1, \ldots, f_n$ are not $\fnull$ and distinct, 
and $\psi$ contains no function definitions.
We will commonly call $\psi$ the \emph{conjecture} of $\varphi$.
\ct{Don't we need additional restrictions on the $\varphi_i$'s here?
(E.g., they are quantifier-free, or perhaps existential or containing only let binders.)}
\end{definition}

Given a signature $\Sigma$, 
we will commonly parition its function symbols $\sfuns{\Sigma}$ into $\sfundefs{\Sigma} \mathrel{\uplus} \sfunndefs{\Sigma}$,
where $\sfundefs{\Sigma}$ is the set of \emph{defined} function symbols,
and consider $\Sigma$-formulas that are well-defined with respect to functions $\sfundefs{\Sigma}$.

A set of function definitions $\{ \varphi_1, \ldots, \varphi_n \}$ 
is \emph{well-specified (in $T$)} if $\varphi_1 \wedge \ldots \wedge \varphi_n$ is $T$-satisfiable.
For instance, $\{ \forallf{\con{f}} \vec x.\;\, \con{f}( x ) \teq 0 \}$ is well-specified,
whereas $\{ \forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ) + 1 \}$ is not. 
Note that a well-specified function definition need not specify a terminating function, e.g.
$\{ \forallf{\con{f}} x^\sortint.\;\, \con{f}( x ) \teq \con{f}( x - 1 ) \}$ or even $\{ \forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ) \}$ are well-specified.
\ajr{revise}

@@

Recent techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} 
have focused on finding models for quantified formulas in SMT.
A instantiation-based procedure is shown in~\cite{GeDeM-CAV-09} to be complete for formulas in the \emph{essentially uninterpreted} fragment.
This fragment is limited to universally quantified formulas where all occurrences of variables occur as direct subchildren of uninterpreted functions.
For instance, the formula $\forall x\typ{\sortint}.\;\, \con{f}( x ) \teq g( x ) + 5$ 
is in the essentially uninterpreted fragment since all occurrences of $x$ occur beneath uninterpreted functions $\con{f}$ and $g$.
Other syntactic criteria are identified in this paper that extend this fragment slightly, including cases when variables occur as arguments of arithmetic predicates.
The techniques in~\cite{ReyEtAl-1-RR-13} can be used for finding finite models for quantified formulas over uninterpreted types as well as types having a 
fixed finite interpretation (such as fixed-width bitvectors, datatype enumerations, etc.).
For instance, 
finite model finding techniques can find a model for
the formula $\forall x\typ{\tau} y\typ{\tau}.\;\, \neg x \teq y \vee \con{f}( x ) \teq \con{f}( y )$
where the uninterpreted type $\tau$ is interpreted as a finite set, and $\con{f}$ as injective finite map.

@@

In the remainder of the section, we fix a signature $\Sigma$.
Say that $\Sigma$-formula $\varphi$ is well-defined with respect to functions $\sfundefs{\Sigma}$, 
and the set of function definitions corresponding to $\sfundefs{\Sigma}$ are well-specified.
%We may translate $\varphi$ into a equisatisfiable formula $\psi$ for which known model-finding procedures~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.
This section presents a method for constructing an extended signature $\extendsig{ \Sigma }$,
and a $\extendsig{ \Sigma }$-formula $\psi$ such that $\varphi$ and $\psi$ are equisatisfiable.
The intuition behind this translation is to use an uninterpreted type \ct{to be defined} to abstract the set of \emph{relevant} tuples for each recursive function $\con{f}$,
and restrict the quantification of the function definition for $\con{f}$ to a variable of this type.
Informally, the relevant tuples $\vec t$ of a function $\con{f}$ are the ones for which the interpretation of $\con{f}( \vec t )$ is relevant to the satisfiability of $\varphi$.

\begin{figure}[t]
\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv( \varphi )$ : 
 \begin{itemize}
   \item[] $\mathsf{let}$ $( \psi, D ) = \conv( \varphi, \ptrue )$;
   %\item[] $\mathsf{assert}( D = \emptyset )$;
   \item[] $\mathsf{return}$ $\psi$
  \end{itemize}
\end{enumerate}
\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv_0( t\typ{\tau}, p )$ : 
 \begin{itemize}
   \item[] $\mathsf{if}$ $\tau \equiv \Bool$ and $t \equiv t_1 \boolop \ldots \boolop t_n$ :
    \begin{itemize}
      \item[] $\mathsf{let}$ $( r_i, D_i ) = \conv_0( t_i, \pol( \boolop, i, p ) )$ for $i = 1, \ldots, n$;
      \item[] $\mathsf{if}$ $p = \ptrue$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \wedge D_1 \wedge \ldots \wedge D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$ $\mathsf{if}$ $p = \pfalse$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \vee \neg D_1 \vee \ldots \vee \neg D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( r_1 \boolop \ldots \boolop r_n, D_1 \cup \ldots \cup D_n )$
      \end{itemize}
    \end{itemize}
  \item[] $\mathsf{else}$ $\mathsf{if}$ $t \equiv \forallf{\con{f}} \vec x.\;\, t_1$
    \begin{itemize}
      \item[] $\mathsf{let} ( r_1, D_1 ) = \conv_0( t_1, p )$;
      \item[] $\mathsf{if}$ $\con{f} \in \sfundefs{\Sigma}$
      \begin{itemize}
        %\item[] $\mathsf{assert}( D_1 = \emptyset )$;
        \item[] $\mathsf{return}$ $( \forall y\typ{\fargsort{\con{f}}}.\;\, ( r_1 [ \vecfarg{\con{f}}( y ) / \vec x ] ), \emptyset )$
      \end{itemize}
      \item[] else
      \begin{itemize}
        \item[] $\mathsf{return}$ $( \forall \vec x.\; r_1, \forall \vec x.\; D_1 )$
      \end{itemize}
    \end{itemize}
   \item[] $\mathsf{else}$
   \begin{itemize}
     \item[] $\mathsf{return}$ $( t, \{ \exists z\typ{\fargsort{\con{f}}}. ( \vecfarg{f}( z ) \teq \vec t ) \mid f( \vec t ) \in \terms( t ), f \in \sfundefs{\Sigma} \} )$
   \end{itemize}
 \end{itemize}
\end{enumerate}
\vspace{-2ex}
\caption{A translation procedure $\conv$ for a $\Sigma$-formula that is well-defined with respect to a set of well-specified function definitions $\sfundefs{\Sigma}$.
In the subprocedure $\conv_0$, the procedure $\pol( \boolop, i, p )$ returns the polarity of the $i^{th}$ child of a $\boolop$-application having polarity $p$,
where $\boolop$ is an interpreted predicate.
}
\label{fig:encoding}
\end{figure}

@@

\begin{example}
\label{ex:translation}
Consider the (combined) signature of linear arithmetic and uninterpreted functions $\Sigma = \Sigmalia \cup \Sigma_{\euf}$,
where $\ssorts{\Sigma_{\euf}} = \{ \sortint \}$ and 
$\sfuns{\Sigma_{\euf}} = \{ s\typ{ \sortint \rightarrow \sortint }, c\typ{ \sortint } \}$.
Consider the $\Sigma$-formula $\varphi$:
\begin{eqnarray} \label{eq:ex-before}
\forall_{s} x\typ{\sortint}.\; \lite( x \leq 0, s(x) \teq 0, s( x ) \teq x + s( x-1 ) ) \wedge s( c ) > 100
\end{eqnarray} 
Here, $s$ is a function that returns the sum of positive numbers between $0$ and its argument $x$.
The formula $\varphi$ is well-defined with respect to functions $\sfundefs{\Sigma_u}$,
and states that the sum of all positive numbers between $0$ and $c$ is greater than $100$.
This formula is satisfiable with a model that interprets $c$ as an integer greater than or equal to $14$.
Due to the universally quantified formula in the left disjunct,
current SMT techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are unable to find a model for this formula.
The signature $\extendsig{\Sigma}$ includes the type $\fargsort{s}$,
and the uninterpreted function $\farg{s}$ of type $\fargsort{s} \rightarrow \sortint$.
The result of $\conv( \varphi, \ptrue )$ is the pair $( \psi, \emptyset )$, where after simplification, $\psi$ is the $\extendsig{\Sigma}$-formula:
\begin{eqnarray} \label{eq:ex-after}
\begin{split}
\forall y\typ{\fargsort{s}}.\;\, \lite( & \farg{s}( y ) \leq 0,  \\
 & s(\farg{s}( y )) \teq 0, \\
 & s(\farg{s}( y )) \teq \farg{s}( y ) + s( \farg{s}( y )-1 ) ) \\
 & \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq \farg{s}( y )-1 )
\end{split}  
\wedge ( s( c ) > 100 \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq c )
\end{eqnarray} 
The universal quantification in the resulting formula is over an uninterpreted type $\fargsort{s}$, 
thus enabling finite model finding techniques~\cite{ReyEtAl-1-RR-13} to search for a finite interpretation for $\fargsort{s}$.
Moreover, all occurrences of $y$ in this formula are beneath applications of the uninterpreted function $\farg{s}$,
implying that the formula is in the essentially uninterpreted fragment of~\cite{GeDeM-CAV-09}, for which there exists a complete instantiation procedure.
We observe that both \cvc and \ziii run indefinitely on formula~(\ref{eq:ex-before}),
and both produce a model for formula~(\ref{eq:ex-after}) in $<0.1$ second.
$\square$
\end{example}

It is important to note that the translation $\conv$ does not preserve the models of $\varphi$.
One model $\I$ for formula~(\ref{eq:ex-after}) in the previous example interprets 
$\fargsort{s}$ as a finite set of elements $\{ u_0, \ldots, u_{14} \}$,
$\farg{s}$ as a finite map mapping containing $u_i \mapsto i$ for $i = 0, \ldots, 14$,
$c$ as $14$, 
and $s$ as the (almost constant) function:
\begin{eqnarray}
\lambda x\typ{Int}. \lite( x \teq 0, 0, \lite( x \teq 1, 1, \lite( x \teq 2, 3, \ldots, \lite( x \teq 13, 91, 105 ) \ldots )
\end{eqnarray}
In other words, $s$ is a function mapping $x$ to the sum of all positive integers between $0$ and $x$ when $0 \leq x \leq 13$,
and $105$ otherwise.
Notice that $\I$ is \emph{not} a model for the original formula~(\ref{eq:ex-before}),
since $\I$ interprets e.g.\ $s( -1 )$ and $s( 15 )$ as $105$.
However, under the assumption that function definitions in $\sfundefs{\Sigma}$ are well-specified, 
we claim that $\conv(\varphi)$ is equisatisfiable to $\varphi$ for any input $\varphi$.
The intuition is that the interpretation of a term such as $s( -1 )$ in the previous example is not relevant to the satisfiability of formula~(\ref{eq:ex-before}).
Thus, a satisfiable (resp. unsatisfiable) response from an SMT solver on input $\conv( \varphi )$ implies the existence (resp.\ nonexistence) of a model for $\varphi$.
Moreover, information in models for $\conv( \varphi )$ contains relevant information regarding the models of $\varphi$.
For instance, the model $\I$ for formula~(\ref{eq:ex-after}) described above interprets $c$ as $14$, 
while there exists a model of formula~(\ref{eq:ex-before}) also interprets $c$ as $14$.
In fact, we claim that there exists a model of $\varphi$ that coincides with each model of $\conv( \varphi )$ on its interpretation of all functions in $\sfuns{\Sigma}$ 
other than those in $\sfundefs{\Sigma}$, i.e. $\{ s \}$ in the previous example.
From a practical point of view, this is not an issue, because
$s$ is the very function that was explicitly defined by the user,
and hence already has an intended interpretation.

@@

\begin{theorem}
If $\varphi$ be a $\Sigma$-formula that is well-defined with respect to functions $\sfundefs{\Sigma}$,
and the set of function definitions corresponding to functions in $\sfundefs{\Sigma}$ is well-specified,
then $\varphi$ and $\conv( \varphi )$ are equisatisfiable.
\end{theorem}
\begin{proof}
(Sketch)  First, we show that if $\varphi$ has a model $\I$, then $\conv( \varphi )$ has a model $\J$.
Let $\J$ be the $\extendsig{\Sigma}$-interpretation that interprets all types $\tau \in \ssorts{\Sigma}$ as $\tau^\I$,
all functions $\con{f} \in \sfuns{\Sigma}$ as $\con{f}^\I$,
and for each function $\con{f}$ of type $\tau_1 \times \ldots \tau_n \rightarrow \tau$ in $\sfundefs{\Sigma}$, 
interprets $\fargsort{\con{f}}$ as a tuple of values of type $\tau_1^\I \times \ldots \times \tau_n^\I$,
and each $\farg{\con{f}}^i$ as the $i^{th}$ projection on such tuples for each $i = 1, \ldots, n$.
Note that $\J$ satisfies \emph{every} constraint of the form $\exists z\typ{\fargsort{\con{f}}}. ( \vecfarg{\con{f}}( z ) \teq \vec t )$,
since by our construction of $\J$ there exists a $z$ such that $z^\J = \vec t^\J$.
By case analysis it follows that $\conv_0$ maintains the invariant that
if $\conv_0( t\typ{\Bool}, \ptrue )$ returns $( r\typ{\Bool}, D )$,
then $\I$ satisfies $t$ if and only if $\J$ satisfies $r$.
Thus, we conclude $\J$ is a model of $\conv( \varphi )$.

@@

In total, our benchmark set consisted of 885 benchmarks (640 mutants plus the original 245).
We considered these 885 benchmarks both before and after the translation $\conv$ as described in Section~\ref{sec:encoding}.
We will denote the latter encoding using the suffix {\bf w}.
For instance {\bf isa+w}, contains 79 benchmarks that are the results of applying $\conv$ to each of the 79 benchmarks in {\bf isa}.

@@

The results are shown in Figure~\ref{fig:results}.
The results are divided into benchmarks before and after the translation $\conv$.
First, we note that in no cases did a configuration answer differently across the two encodings.
In other words, there were no cases in which \ziii or \cvc answered e.g.\ ``unsatisfiable" on a benchmark $\varphi$
and ``satisfiable" on its corresponding benchmark $\conv( \varphi )$, or vice versa.
This provides confirming evidence that the translation $\conv$ is \emph{refutation-sound},
as we did not find a case where $\conv( \varphi )$ was unsatisfiable when $\varphi$ was satisfiable.
It also provides confirming evidence that the translation is \emph{model-sound} for this set of benchmarks,
as we did not find a case where $\conv( \varphi )$ was satisfiable when $\varphi$ was unsatisfiable.
This is to be expected, since the functions in our benchmark sets are expected to be well-specified (and in fact, terminating).

@@

We conclude from this evaluation that the translation $\conv$ enables SMT solvers to find counterexamples 
for conjectures involving recursive functions whose definitions are well-specified.
Both {\bf z3} and {\bf cvc4+f} were able to find a large number of counterexamples after the translation,
with the latter configuration finding significantly more instances (331 to 112).

@@

We have seen a satisfiability-preserving translation $\conv$ that significantly improves the effectiveness of known SMT techniques
for finding counterexamples for formulas involving recursive function definitions.
For convienience, the translation $\conv$ has been implemented in the latest development version of \cvc~\cite{ReyEtAl-1-RR-13} (version 1.5 pre-release).
Function definitions $\forallf{\con{f}} \vec x.\; \varphi$ can be written in using the $\definefunreccmd$ command, 
from Cersion 2.5~\cite{} of the SMT-LIB standard.
For instance, the formula $\varphi$ in Example~\ref{ex:translation} 
can be written as:

>>>>

%In this paper, we associate quantified formulas with uninterpreted functions (informally, those whose semantics they define).
In this paper, we consider \emph{annotated quantifed formulas} of the form
$\forallf{a} \vec x.\; \varphi$, where the annotation~$a$ is either
an uninterpreted function $\con{f} \in \sfuns{\Sigma}$ or the special symbol
$\fnull$.
Given $\con{f} : \tau_1 \times \ldots \times \tau_n \rightarrow \tau$,
an annotated
quantified formula $\forallf{\con{f}} \vec x.\; \varphi$ is a \emph{function definition}
(\,\emph{for $\con{f}$}\vthinspace) if $\vec x$ is a tuple of variables of type
$\tau_1, \ldots, \tau_n$ and $\forallf{\con{f}} \vec x.\; \varphi$ is equivalent to
$\forall \vec x.\;\, \con{f}( \vec x ) \teq t$ for some $t$. We write $\forall \vec
x.\; \varphi$ as shorthand for $\forallf{\fnull} \vec x.\; \varphi$, and $\exists
\vec x. \varphi$ as shorthand for $\neg \forall \vec x.\; \neg \varphi$.

\begin{definition}%
\rm
A formula $\varphi$ is in \emph{definitional form with respect to} functions $\{
\con{f}_1, \ldots, \con{f}_n \}$ if it is of the form
%
%\kern-\abovedisplayskip
%\kern+\abovedisplayshortskip
%
\begin{equation} \label{eq:wdf}
(\forallf{\con{f}_1} \vec x_1.\; \varphi_1) \wedge \ldots \wedge (\forallf{\con{f}_n} \vec x_n.\; \varphi_n) \wedge \psi
\end{equation}
%
where $\con{f}_1, \ldots, \con{f}_n$ are not $\fnull$ and distinct, 
and $\psi$ contains no function definitions.
We call $\psi$ the \emph{conjecture} of $\varphi$.
\ct{Don't we need additional restrictions on the $\varphi_i$'s here?
(E.g., they are quantifier-free, or perhaps existential or containing only let binders.)}
\end{definition}

Given a signature $\Sigma$,
we partition its function symbols $\sfuns{\Sigma}$ into $\sfundefs{\Sigma} \mathrel{\uplus} \sfunndefs{\Sigma}$,
where $\sfundefs{\Sigma}$ is the set of \emph{defined} function symbols,
and consider $\Sigma$-formulas that are in definitional form with respect to functions $\sfundefs{\Sigma}$.

A set of function definitions $\{ \varphi_1, \ldots, \varphi_n \}$ 
is \emph{admissible} (\emph{in $T$}) if $\varphi_1 \wedge \ldots \wedge \varphi_n$ is $T$-satisfiable.
For example, $\{ \forallf{\con{f}} \vec x.\;\, \con{f}( x ) \teq 0 \}$ is admissible,
whereas $\{ \forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ) + 1 \}$ is not. 
Note that an admissible function definition need not specify a terminating function, e.g.
$\{ \forallf{\con{f}} x^\sortint.\;\, \con{f}( x ) \teq \con{f}( x - 1 ) \}$ or even $\{ \forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ) \}$ are admissible.
\ajr{revise}

@@

Recent techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} 
have focused on finding models for quantified formulas in SMT.
A instantiation-based procedure is shown in~\cite{GeDeM-CAV-09} to be complete for formulas in the \emph{essentially uninterpreted} fragment.
This fragment is limited to universally quantified formulas where all occurrences of variables occur as direct subchildren of uninterpreted functions.
For example, the formula $\forall x\typ{\sortint}.\;\, \con{f}( x ) \teq g( x ) + 5$ 
is in the essentially uninterpreted fragment since all occurrences of $x$ occur beneath uninterpreted functions $\con{f}$ and $g$.
Other syntactic criteria are identified in this paper that extend this fragment slightly, including cases when variables occur as arguments of arithmetic predicates.
The techniques in~\cite{ReyEtAl-1-RR-13} can be used for finding finite models for quantified formulas over uninterpreted types as well as types having a 
fixed finite interpretation (such as fixed-width bitvectors, datatype enumerations, etc.).
For example, finite model finding techniques can find a model for
the formula $\forall x\typ{\tau} y\typ{\tau}.\;\, \neg x \teq y \vee \con{f}( x ) \teq \con{f}( y )$
where the uninterpreted type $\tau$ is interpreted as a finite set, and $\con{f}$ as injective finite map.

@@

In the remainder of the section, we fix a signature $\Sigma$.
Say that $\Sigma$-formula $\varphi$ is in definitional form with respect to functions $\sfundefs{\Sigma}$, 
and the set of function definitions corresponding to $\sfundefs{\Sigma}$ are admissible.
%We may translate $\varphi$ into a equisatisfiable formula $\psi$ for which known model-finding procedures~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.
This section presents a method for constructing an extended signature $\extendsig{ \Sigma }$,
and a $\extendsig{ \Sigma }$-formula $\psi$ such that $\varphi$ and $\psi$ are equisatisfiable.
The intuition behind this translation is to use an uninterpreted type \ct{to be defined} to abstract the set of \emph{relevant} tuples for each recursive function $\con{f}$,
and restrict the quantification of the function definition for $\con{f}$ to a variable of this type.
Informally, the relevant tuples $\vec t$ of a function $\con{f}$ are the ones for which the interpretation of $\con{f}( \vec t )$ is relevant to the satisfiability of $\varphi$.

\begin{figure}[t]
\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv( \varphi )$ : 
 \begin{itemize}
   \item[] $\mathsf{let}$ $( \psi, D ) = \conv( \varphi, \ptrue )$;
   %\item[] $\mathsf{assert}( D = \emptyset )$;
   \item[] $\mathsf{return}$ $\psi$
  \end{itemize}
\end{enumerate}
\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv_0( t\typ{\tau}, p )$ : 
 \begin{itemize}
   \item[] $\mathsf{if}$ $\tau \equiv \Bool$ and $t \equiv t_1 \boolop \ldots \boolop t_n$ :
    \begin{itemize}
      \item[] $\mathsf{let}$ $( r_i, D_i ) = \conv_0( t_i, \pol( \boolop, i, p ) )$ for $i = 1, \ldots, n$;
      \item[] $\mathsf{if}$ $p = \ptrue$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \wedge D_1 \wedge \ldots \wedge D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$ $\mathsf{if}$ $p = \pfalse$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \vee \neg D_1 \vee \ldots \vee \neg D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( r_1 \boolop \ldots \boolop r_n, D_1 \cup \ldots \cup D_n )$
      \end{itemize}
    \end{itemize}
  \item[] $\mathsf{else}$ $\mathsf{if}$ $t \equiv \forallf{\con{f}} \vec x.\;\, t_1$
    \begin{itemize}
      \item[] $\mathsf{let} ( r_1, D_1 ) = \conv_0( t_1, p )$;
      \item[] $\mathsf{if}$ $\con{f} \in \sfundefs{\Sigma}$
      \begin{itemize}
        %\item[] $\mathsf{assert}( D_1 = \emptyset )$;
        \item[] $\mathsf{return}$ $( \forall y\typ{\fargsort{\con{f}}}.\;\, ( r_1 [ \vecfarg{\con{f}}( y ) / \vec x ] ), \emptyset )$
      \end{itemize}
      \item[] else
      \begin{itemize}
        \item[] $\mathsf{return}$ $( \forall \vec x.\; r_1, \forall \vec x.\; D_1 )$
      \end{itemize}
    \end{itemize}
   \item[] $\mathsf{else}$
   \begin{itemize}
     \item[] $\mathsf{return}$ $( t, \{ \exists z\typ{\fargsort{\con{f}}}. ( \vecfarg{f}( z ) \teq \vec t ) \mid f( \vec t ) \in \terms( t ), f \in \sfundefs{\Sigma} \} )$
   \end{itemize}
 \end{itemize}
\end{enumerate}
\vspace{-2ex}
\caption{A translation procedure $\conv$ for a $\Sigma$-formula that is in definitional form with respect to a set of admissible function definitions $\sfundefs{\Sigma}$.
In the subprocedure $\conv_0$, the procedure $\pol( \boolop, i, p )$ returns the polarity of the $i^{th}$ child of a $\boolop$-application having polarity $p$,
where $\boolop$ is an interpreted predicate.
}
\label{fig:encoding}
\end{figure}

@@

\begin{example}
\label{ex:translation}
Consider the (combined) signature of linear arithmetic and uninterpreted functions $\Sigma = \Sigmalia \cup \Sigma_{\euf}$,
where $\ssorts{\Sigma_{\euf}} = \{ \sortint \}$ and 
$\sfuns{\Sigma_{\euf}} = \{ s\typ{ \sortint \rightarrow \sortint }, c\typ{ \sortint } \}$.
Consider the $\Sigma$-formula $\varphi$:
\begin{equation} \label{eq:ex-before}
\forall_{s} x\typ{\sortint}.\; \lite( x \leq 0, s(x) \teq 0, s( x ) \teq x + s( x-1 ) ) \wedge s( c ) > 100
\end{equation} 
Here, $s$ is a function that returns the sum of positive numbers between $0$ and its argument $x$.
The formula $\varphi$ is in definitional form with respect to functions $\sfundefs{\Sigma_u}$,
and states that the sum of all positive numbers between $0$ and $c$ is greater than $100$.
This formula is satisfiable with a model that interprets $c$ as an integer greater than or equal to $14$.
Due to the universally quantified formula in the left disjunct,
current SMT techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are unable to find a model for this formula.
The signature $\extendsig{\Sigma}$ includes the type $\fargsort{s}$,
and the uninterpreted function $\farg{s}$ of type $\fargsort{s} \rightarrow \sortint$.
The result of $\conv( \varphi, \ptrue )$ is the pair $( \psi, \emptyset )$, where after simplification, $\psi$ is the $\extendsig{\Sigma}$-formula:
\begin{eqnarray} \label{eq:ex-after}
\begin{split}
\forall y\typ{\fargsort{s}}.\;\, \lite( & \farg{s}( y ) \leq 0,  \\
 & s(\farg{s}( y )) \teq 0, \\
 & s(\farg{s}( y )) \teq \farg{s}( y ) + s( \farg{s}( y )-1 ) ) \\
 & \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq \farg{s}( y )-1 )
\end{split}  
\wedge ( s( c ) > 100 \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq c )
\end{eqnarray} 
The universal quantification in the resulting formula is over an uninterpreted type $\fargsort{s}$, 
thus enabling finite model finding techniques~\cite{ReyEtAl-1-RR-13} to search for a finite interpretation for $\fargsort{s}$.
Moreover, all occurrences of $y$ in this formula are beneath applications of the uninterpreted function $\farg{s}$,
implying that the formula is in the essentially uninterpreted fragment of~\cite{GeDeM-CAV-09}, for which there exists a complete instantiation procedure.
We observe that both \cvc and \ziii run indefinitely on formula~(\ref{eq:ex-before}),
and both produce a model for formula~(\ref{eq:ex-after}) in $<0.1$ second.
$\square$
\end{example}

It is important to note that the translation $\conv$ does not preserve the models of $\varphi$.
One model $\I$ for formula~(\ref{eq:ex-after}) in the previous example interprets 
$\fargsort{s}$ as a finite set of elements $\{ u_0, \ldots, u_{14} \}$,
$\farg{s}$ as a finite map mapping containing $u_i \mapsto i$ for $i = 0, \ldots, 14$,
$c$ as $14$, 
and $s$ as the (almost constant) function:
\begin{equation}
\lambda x\typ{Int}. \lite( x \teq 0, 0, \lite( x \teq 1, 1, \lite( x \teq 2, 3, \ldots, \lite( x \teq 13, 91, 105 ) \ldots )
\end{equation}
In other words, $s$ is a function mapping $x$ to the sum of all positive integers between $0$ and $x$ when $0 \leq x \leq 13$,
and $105$ otherwise.
Notice that $\I$ is \emph{not} a model for the original formula~(\ref{eq:ex-before}),
since $\I$ interprets e.g.\ $s( -1 )$ and $s( 15 )$ as $105$.
However, under the assumption that function definitions in $\sfundefs{\Sigma}$ are admissible, 
we claim that $\conv(\varphi)$ is equisatisfiable to $\varphi$ for any input $\varphi$.
The intuition is that the interpretation of a term such as $s( -1 )$ in the previous example is not relevant to the satisfiability of formula~(\ref{eq:ex-before}).
Thus, a satisfiable (resp. unsatisfiable) response from an SMT solver on input $\conv( \varphi )$ implies the existence (resp.\ nonexistence) of a model for $\varphi$.
Moreover, information in models for $\conv( \varphi )$ contains relevant information regarding the models of $\varphi$.
For example, the model $\I$ for formula~(\ref{eq:ex-after}) described above interprets $c$ as $14$, 
while there exists a model of formula~(\ref{eq:ex-before}) also interprets $c$ as $14$.
In fact, we claim that there exists a model of $\varphi$ that coincides with each model of $\conv( \varphi )$ on its interpretation of all functions in $\sfuns{\Sigma}$ 
other than those in $\sfundefs{\Sigma}$, i.e. $\{ s \}$ in the previous example.
From a practical point of view, this is not an issue, because
$s$ is the very function that was explicitly defined by the user,
and hence already has an intended interpretation.

@@

\begin{theorem}
If $\varphi$ be a $\Sigma$-formula that is in definitional form with respect to functions $\sfundefs{\Sigma}$,
and the set of function definitions corresponding to functions in $\sfundefs{\Sigma}$ is admissible,
then $\varphi$ and $\conv( \varphi )$ are equisatisfiable.
\end{theorem}
\begin{proof}
(Sketch)  First, we show that if $\varphi$ has a model $\I$, then $\conv( \varphi )$ has a model $\J$.
Let $\J$ be the $\extendsig{\Sigma}$-interpretation that interprets all types $\tau \in \ssorts{\Sigma}$ as $\tau^\I$,
all functions $\con{f} \in \sfuns{\Sigma}$ as $\con{f}^\I$,
and for each function $\con{f}$ of type $\tau_1 \times \ldots \tau_n \rightarrow \tau$ in $\sfundefs{\Sigma}$, 
interprets $\fargsort{\con{f}}$ as a tuple of values of type $\tau_1^\I \times \ldots \times \tau_n^\I$,
and each $\farg{\con{f}}^i$ as the $i^{th}$ projection on such tuples for each $i = 1, \ldots, n$.
Note that $\J$ satisfies \emph{every} constraint of the form $\exists z\typ{\fargsort{\con{f}}}. ( \vecfarg{\con{f}}( z ) \teq \vec t )$,
since by our construction of $\J$ there exists a $z$ such that $z^\J = \vec t^\J$.
By case analysis it follows that $\conv_0$ maintains the invariant that
if $\conv_0( t\typ{\Bool}, \ptrue )$ returns $( r\typ{\Bool}, D )$,
then $\I$ satisfies $t$ if and only if $\J$ satisfies $r$.
Thus, we conclude $\J$ is a model of $\conv( \varphi )$.

@@

In total, our benchmark set consisted of 885 benchmarks (640 mutants plus the original 245).
We considered these 885 benchmarks both before and after the translation $\conv$ as described in Section~\ref{sec:encoding}.
We will denote the latter encoding using the suffix {\bf w}.
For example, {\bf isa+w} contains 79 benchmarks that are the results of
applying $\conv$ to each of the 79 benchmarks in {\bf isa}.

@@

The results are shown in Figure~\ref{fig:results}.
The results are divided into benchmarks before and after the translation $\conv$.
First, we note that in no cases did a configuration answer differently across the two encodings.
In other words, there were no cases in which \ziii or \cvc answered e.g.\ ``unsatisfiable" on a benchmark $\varphi$
and ``satisfiable" on its corresponding benchmark $\conv( \varphi )$, or vice versa.
This provides confirming evidence that the translation $\conv$ is \emph{refutation-sound},
as we did not find a case where $\conv( \varphi )$ was unsatisfiable when $\varphi$ was satisfiable.
It also provides confirming evidence that the translation is \emph{model-sound} for this set of benchmarks,
as we did not find a case where $\conv( \varphi )$ was satisfiable when $\varphi$ was unsatisfiable.
This is to be expected, since the functions in our benchmark sets are expected to be admissible (and in fact, terminating).

@@

We conclude from this evaluation that the translation $\conv$ enables SMT solvers to find counterexamples 
for conjectures involving recursive functions whose definitions are admissible.
Both {\bf z3} and {\bf cvc4+f} were able to find a large number of counterexamples after the translation,
with the latter configuration finding significantly more instances (331 to 112).

@@

We have seen a satisfiability-preserving translation $\conv$ that significantly improves the effectiveness of known SMT techniques
for finding counterexamples for formulas involving recursive function definitions.
For convienience, the translation $\conv$ has been implemented in the latest development version of \cvc~\cite{ReyEtAl-1-RR-13} (version 1.5 pre-release).
Function definitions $\forallf{\con{f}} \vec x.\; \varphi$ can be written in using the $\definefunreccmd$ command, 
from Version 2.5~\cite{} of the SMT-LIB standard.
For example, the formula $\varphi$ in Example~\ref{ex:translation} 
can be written as:
